Inferring cosmology from gravitational waves using non-parametric detector-frame mass distribution

### 2024-10-16 src/tex/ms.tex
[showyourwork] overleaf sync

The fundamental idea of the spectral-siren method is to infer cosmology by comparing the observed mass population distribution with the intrinsic mass population distribution since the observed distribution is redshifted due to the expansion of the Universe.
The method works best when the intrinsic distribution is well understood.
However, the intrinsic distribution is not known, and the method depends on the choice of the population model.
Furthermore, the intrinsic distribution is degenerate with cosmology: for example, an intrinsic distribution with a linear redshift evolution is degenerate with $H_0$, as they both redshift the observed distribution in the same way.
While this example is not physical, this degeneracy still affects the inference of $H_0$.
Due to these issues, the spectral-siren method is not straightforward to apply, intrinsic population model parameters and cosmological parameters are required to be inferred simultaneously, which makes the method computationally expensive.

An alternative approach to the spectral-siren method is to use non-parametric methods to reconstruct the observed mass population distribution directly from the data.
Non-parametric methods allow us to reconstruct arbitrary probability densities without making assumptions about the form of the distribution \citep{Rinaldi:2021bhm}.
In our context, we can use non-parametric methods to reconstruct the observed mass population distribution directly from the data.
We can then compare the reconstructed distribution with the predicted distribution from some intrinsic population model and cosmological model, which allows us to infer the model parameters by minimizing the distance between the two distributions.

@@

With these modifications, we separate the spectral-siren method into two distinct parts.
The first part is the reconstruction of the observed population from the data, which does not require any assumptions about the population model, and the intermediate result is completely data-driven.
We make use of the \ac{(H)DPGMM} developed in \cite{Rinaldi:2021bhm} as a non-parametric model to reconstruct the observed population distribution.
Other assumptions including the population model, cosmological model and selection function are then incorporated in the second half, where we transform the source-frame population model to the detector frame with a given cosmological model.
This separation allows us to explore the impact of different assumptions easily.
The intermediate result of the first part can be reused as long as the data does not change, reducing the computational cost of the method.

@@

The selection effect is chosen to be considered in the transformation instead of the reconstruction, as the uncertainty in the reconstruction will diverge in the region where the selection function censoring the data.
For example, if we consider the selection function in the reconstruction, the uncertainty in the reconstruction will be large at the low mass end, as the selection function censors the low mass events effectively.
\cite{Essick:2023upv} also pointed out that including the selection function in the reconstruction may lead to insignificant biases in the analysis.
Therefore, we consider the selection function in the transformation to avoid these issues.

@@

The optimization procedure does not scale with the number of \ac{GW} events, as the optimization is performed for each reconstructed distribution.
Therefore, the more \ac{GW} events we have, the longer the runtime of the reconstruction, but the runtime of the optimization remains the same.
This is advantageous for performing the analysis with different models when more \ac{GW} events are available, as the reconstruction can be reused.
Other than that, the optimization procedure can be performed in parallel for each reconstructed distribution, as each reconstructed distribution is independent.
This reduces the runtime as long as computational resources are available.

@@

After the transformation, we applied a selection function to the $m^z_1$ and $d_L$ samples.
We used the physically modelled selection function implemented in \cite{Lorenzo-Medina:2024opt}.
Note that we marginalized over $q$ for the selection function, similar to the transformation in Sec.~\ref{sec:transformation}.
After applying the selection function, there were \variable{output/n_obs_samples.txt} sets of samples left.

@@

The result of the \ac{PE} is shown in Fig.~\ref{fig:simulation_result_H0}.
\begin{figure}[htbp]
    \script{plot_simulation_result_H0.py}
    \includegraphics[width=\linewidth]{figures/simulation_result_H0.pdf}
    \caption{
        The histogram shows the distribution of the optimized $H_0$.
        The vertical blue line and shaded regions represent the median, 68\%, and 90\% credible intervals of the distribution.
        The vertical orange line represents the true $H_0$ we used to generate the mock data.
    }
    \label{fig:simulation_result_H0}
\end{figure}
We find that the $H_0$ distribution, whose associated uncertainty comes from the reconstruction uncertainty, is consistent with the simulated $H_0$ value.

@@

In this subsection, we test our method with a more realistic case where we infer $H_0$ and a subset of the population model parameters.
Similar to the previous subsection, we assume perfect knowledge of both models, the selection function and all the parameters that are not objects of the inference.
Other than $H_0$, we also chose the power law index $\alpha$, the peak mass $\mu$, and the peak width $\sigma$ of the \textsc{power law + peak} model as the free parameters.
The minimum and maximum mass, and the range of mass tapering at the low mass end $\delta$ are fixed, as we may not be able to infer these parameters accurately with our method.
This is because the reconstructions are only valid in the region where the samples are present, and they may be not accurate at the boundaries depending on the boundary features of the observed population distribution.
In principle, more parameters can be inferred with our method, for example, the peak weight $w$ in the mass population model and the present mass density $\Omega_m$ in the $\Lambda$CDM model.
However, the optimization procedure becomes more computationally expensive as the number of parameters increases, therefore, we only chose a subset of the parameters to demonstrate the idea of our method.

@@

In this example, instead of transforming the source-frame population model to the detector frame for each $H_0$ and performing a grid search, we define a function that calculates $d_\mathrm{JS}$ for a given set of the free parameters.
We then use the optimization algorithm mentioned in Sec.~\ref{sec:pe} to find the optimized free parameters that minimize $d_\mathrm{JS}$.
The optimization method requires bounds for the free parameters, which we show in Table~\ref{tab:bounds}.
\begin{table}[htbp]
    \caption{Bounds for the free parameters}
    \begin{tabular}{cc}
        \toprule
        Parameter & Bounds \\
        \midrule
        $H_0$ & $[10, 300]$ \\
        $\alpha$ & $[1.01, 10]$ \\
        $\mu$ & $[10, 70]$ \\
        $\sigma$ & $[0.01, 10]$ \\
        \botrule
    \end{tabular}
    \label{tab:bounds}
\end{table}
The initial guesses for the optimization are chosen randomly from a uniform distribution within the bounds.
The result of the \ac{PE} is shown in Fig.~\ref{fig:simulation_result_multi}.
\begin{figure}[htbp]
    \script{plot_simulation_result_multi.py}
    \includegraphics[width=\linewidth]{figures/simulation_result_multi.pdf}
    \caption{
        The corner plot shows the distribution of the optimized parameters.
        The diagonal histograms show the marginalized distributions.
        The off-diagonal contour plots show the joint distributions, where the contours represent the 39.35\%, 90\% credible intervals.
        The red lines and points represent the true values we used to generate the mock data.
    }
    \label{fig:simulation_result_multi}
\end{figure}
We find that the distribution, whose associated uncertainty comes from the reconstruction uncertainty, is consistent with the simulated values.
Some features of the result are discussed together with the result of the real data in Sec.~\ref{sec:real_data}.

@@

In this section, we apply our method to the real data from the \ac{LVK}.
We included all $70$ events labeled as BBH in Table~I of \cite{KAGRA:2021duu}.

>>>>

The fundamental idea of the spectral-siren method is to infer cosmology by comparing the observed mass population distribution with the intrinsic mass population distribution since the observed distribution is redshifted due to the expansion of the Universe.
The method works best when the intrinsic distribution is well understood.
However, the features present in the mass distribution are not so well characterized yet, and their astrophysical origin is a debated topic \citep{Mapelli:2020vfa, Mandel:2018hfr}.
In this case, the spectral-siren method depends on the choice of the population model, which can introduce biases in the inference \citep{Mastrogiovanni:2021wsd, Pierra:2023deu}.
Furthermore, the intrinsic distribution is degenerate with cosmology: for example, an intrinsic distribution with a linear redshift evolution is degenerate with $H_0$, as they both redshift the observed distribution in the same way.
While this example is not physical, this degeneracy still affects the inference of $H_0$.
Due to these issues, astrophysics and cosmology are tightly linked to each other, and a proper inference of the astrophysical and cosmological parameters should be done simultaneously, which makes the method computationally expensive.

A possible alternative approach, which we will propose in this paper, is to use non-parametric methods to reconstruct the observed mass population distribution directly from the data.
Non-parametric methods allow us to reconstruct arbitrary probability densities without making assumptions about the form of the distribution \citep{Rinaldi:2021bhm}.
In our context, we can use non-parametric methods to reconstruct the observed mass population distribution directly from the data.
We can then compare the reconstructed distribution with the predicted distribution from some intrinsic population model and cosmological model, which allows us to infer the model parameters by minimizing the distance between the two distributions.

@@

With these modifications, we separate the spectral-siren method into two distinct parts.
The first part is the reconstruction of the observed population from the data, which does not require any assumptions about the population model, and the intermediate result is completely data-driven.
We make use of the \ac{(H)DPGMM} developed in \cite{Rinaldi:2021bhm} as a non-parametric model to reconstruct the observed population distribution.
In the second part, we make use of all the other assumptions, namely the population model, cosmological model and selection function, to transform the source-frame population model into the detector frame.
This separation allows us to explore the impact of different assumptions easily.
The intermediate result of the first part can be reused as long as the data does not change, reducing the computational cost of the method.

@@

The selection effect is chosen to be considered in the transformation instead of the reconstruction, as the uncertainty in the reconstruction will diverge in the region where the selection function censors the data.
For example, if we consider the selection function in the reconstruction, the uncertainty in the reconstruction will be large at the low mass end, as the selection function censors the low mass events effectively.
This way of including selection effects in the analysis, albeit simple and straightforward, is only an approximation of the real selection process, as pointed out in \cite{Essick:2023upv}.
Nonetheless, the large uncertainties currently associated with \ac{GW} measurements are most likely to be dominant with respect to the bias induced by this approximation.
Therefore, we opted for including the selection function in the transformation itself.
We will explore an exact way of including selection effects in this method in a future paper.
Therefore, we consider the selection function in the transformation to avoid these issues.

@@

The optimization procedure is highly efficient as it does not scale with the number of \ac{GW} events, since the optimization is performed for each reconstructed distribution independently.
Therefore, even as the number of \ac{GW} events increases, the runtime of the optimization remains constant, making our method scalable and robust for future large datasets.
This is advantageous for performing the analysis with different models when more \ac{GW} events are available, as the reconstruction can be reused.
Other than that, the optimization procedure can be performed in parallel for each reconstructed distribution, as each reconstructed distribution is independent.
This reduces the runtime as long as computational resources are available.

@@

After the transformation, we applied a selection function to the $m^z_1$ and $d_L$ samples.
We used the physically modelled selection function implemented in \cite{Lorenzo-Medina:2024opt}.
Note that we marginalized over $q$ for the selection function, similar to the transformation in Sec.~\ref{sec:transformation}.
After applying the selection function, which filters out the non-detectable events based on their $m^z_1$ and $d_L$, we retained \variable{output/n_obs_samples.txt} sets of samples for further analysis.

@@

The result of the \ac{PE} is shown in Fig.~\ref{fig:simulation_result_H0}.
\begin{figure}[htbp]
    \script{plot_simulation_result_H0.py}
    \includegraphics[width=\linewidth]{figures/simulation_result_H0.pdf}
    \caption{
        The histogram shows the distribution of the optimized $H_0$.
        The vertical blue line and shaded regions represent the median, 68\%, and 90\% credible intervals of the distribution.
        The vertical orange line represents the true $H_0$ we used to generate the mock data.
    }
    \label{fig:simulation_result_H0}
\end{figure}
The recovered $H_0$ distribution, whose uncertainty budget comes from the remapping of the uncertainty in the non-parametric reconstruction, is consistent with the simulated value.

@@

In this subsection, we test our method with a more realistic case where we infer $H_0$ and a subset of the population model parameters.
Similar to the previous subsection, we assume perfect knowledge of both models, the selection function and all the parameters that are not objects of the inference.
Other than $H_0$, we included the power law index $\alpha$, the peak mass $\mu$, and the peak width $\sigma$ of the \textsc{power law + peak} model as the free parameters in the analysis.
The minimum and maximum mass, and the range of mass tapering at the low mass end $\delta$ are fixed.
These parameters, which control the boundary features of the observed population distribution, were chosen because the reconstructions are only valid in the region where the samples are present, and the reconstructed distributions may not be accurate at the boundaries depending on the boundary features of the observed population distribution.
In principle, more parameters can be inferred with our method, for example, the peak weight $w$ in the mass population model and the present mass density $\Omega_m$ in the $\Lambda$CDM model.
However, the optimization procedure becomes more computationally expensive as the number of parameters increases, therefore, we only chose a subset of the parameters to demonstrate the idea of our method.

@@

In this example, instead of transforming the source-frame population model to the detector frame for each $H_0$ and performing a grid search, we define a function that calculates $d_\mathrm{JS}$ for a given set of the free parameters.
We then use the optimization algorithm mentioned in Sec.~\ref{sec:pe} to find the optimized free parameters that minimize $d_\mathrm{JS}$.
The optimization method requires bounds for the free parameters, which we show in Table~\ref{tab:bounds}.
\begin{table}[htbp]
    \caption{Bounds for the free parameters}
    \begin{tabular}{cc}
        \toprule
        Parameter & Bounds \\
        \midrule
        $H_0$ & $[10, 300]$ \\
        $\alpha$ & $[1.01, 10]$ \\
        $\mu$ & $[10, 70]$ \\
        $\sigma$ & $[0.01, 10]$ \\
        \botrule
    \end{tabular}
    \label{tab:bounds}
\end{table}
The initial guesses for the optimization are chosen randomly from a uniform distribution within the bounds.
The result of the \ac{PE} is shown in Fig.~\ref{fig:simulation_result_multi}.
\begin{figure}[htbp]
    \script{plot_simulation_result_multi.py}
    \includegraphics[width=\linewidth]{figures/simulation_result_multi.pdf}
    \caption{
        The corner plot shows the distribution of the optimized parameters.
        The diagonal histograms show the marginalized distributions.
        The off-diagonal contour plots show the joint distributions, where the contours represent the 39.35\%, 90\% credible intervals.
        The red lines and points represent the true values we used to generate the mock data.
    }
    \label{fig:simulation_result_multi}
\end{figure}
The recovered distribution, whose uncertainty budget comes from the remapping of the uncertainty in the non-parametric reconstruction, is consistent with the simulated values.
Some features of the result are discussed together with the result of the real data in Sec.~\ref{sec:real_data}.

@@

In this section, we apply our method to the real data from the \ac{LVK}.
We included all $70$ events labeled as BBH in Table~I of \cite{KAGRA:2021duu}, which are different from the events used in \cite{LIGOScientific:2021aug}.
